Supplementary Materials for cILR: Taxonomic enrichment analysis with isometric log ratios

### 2022-01-26 cbea_revised.tex
Updates from Overleaf

@@

However, there exists unique challenges in the analysis of these taxonomic count tables \cite{li2019a,li2015}. The data is sparse, high-dimensional, and likely compositional \cite{gloor2017, li2019a, li2015}. Even though the these problems are challenging, a very approachable solution is to to use set-based analysis methods, also termed gene set testing in the genomics literature \cite{khatri2012, goeman2007}. Aggregated variables can be less sparse, and testing on a smaller number of features can reduce the multiple-testing burden. As such, gene set testing methods have been shown to increase power and reproducibility of genomic analyses. Furthermore, through the usage of functionally informative sets defined apriori based on historical experiments (for example MSigDB \cite{subramanian2005}, and Gene Ontology \cite{ashburner2000}), gene set analyses also allows for more biologically informative interpretations.

There exists a diverse set of methods already developed in this field. Traditional set testing methods utilize the hypergeometric distribution to test for the overrepresentation of a gene set using a candidate list of genes screened based on a marginal model \cite{goeman2007}. Unfortunately, these approaches are sensitive to the differential expression test as well as the chosen threshold where genes can be part of the candidate list. Aggregate score methods, which are generally more preferred \cite{irizarry2009}, instead assigns a score for each gene set based on gene-specific statistics such as z-scores or fold change. Of these approaches, methods such as GSEA \cite{subramanian2005} performs a test for each gene set at the population level, summarizing information across all samples. Conversely, methods such as GSVA \cite{hanzelmann2013} and VAM \cite{frost2020}, generate enrichment scores at the sample level and are more akin to a transformation. In addition to being able to screen for enriched sets per sample, this strategy also allows for the flexible incorporation of different downstream analyses, such as fitting prediction models, or performing dimension reduction.

@@

In microbiome research, even though no explicit enrichment analysis is performed, researchers often involves aggregating taxa to higher Linnean classification levels such as genus, family, or phylum. However, despite this interest, there is limited research done to extend existing set-based methods to microbiome relative abundance data. Some software suites, such as \emph{MicrobiomeAnalyst}, do offer tools to perform enrichment testing with curated taxon sets \cite{chong2020}. However, the approach used in \emph{MicrobiomeAnalyst} is a form of overrepresentation analysis at the population level and therefore similarly sensitive to the differential abundance approach used and p-value threshold. One of the primary challenges for adapting gene set analysis to the microbiome context is the compositional nature of the data. Sequencing technologies constrain the total number of reads, and samples are expected to have the same number of reads instead of DNA content \cite{quinn2019, quinn2018}. However, different samples still yield arbitrarily different total read counts \cite{morton2019, gloor2017}, suggesting the use of normalization methods to allow for proper comparison of feature abundances across samples. However, microbiome data sets do not follow certain assumptions that enable the cross-application of methods from similar fields (such as RNA-seq) \cite{quinn2019, quinn2018}. For example, DESeq2's \emph{estimateSizeFactors} \cite{love2014} assumes that the majority of genes acts as housekeeping genes with constant expression levels across samples. As such, practitioners often rely on total sum normalization to transform count data into relative proportions that sum to one \cite{weiss2017}. Empirical studies have supported this decision, showing increased performance of analysis tasks under this normalization schema \cite{mcknight2019}. Since this approach imposes a sum constraint on the data, post normalization microbiome data sets are compositional \cite{gloor2017}, which means that the abundance of any taxa can only be interpreted relative to another. Under this scenario, log-ratio based approaches from the compositional data analysis (CoDA) literature \cite{aitchison1994} are motivated to address this issue.

@@

The CBEA method generates sample-specific enrichment scores for microbial sets using products of proportions \cite{egozcue2003}. Details on the computational implementation of CBEA can be found in the supplemental materials. The CBEA method takes two inputs:  

@@

This assumption is also supported by preliminary simulation studies (detailed below). In panel A of Fig~\ref{fig:1}, we simulated microbiome taxonomic count data under the global null across different data features and compute raw CBEA scores and compute kurtosis and skewness. It can be seen that the characteristics of the null change depending on sparsity and inter-taxa correlation. Sparsity seems to drive the distribution to be more positively skewed while inter-taxa correlation encourages platykurtic (negative kurtosis). The effect is most dramatic under both high inter-taxa correlation and sparsity. This heterogeneity further supports the decision to estimate an empirical null distribution, similar to Efron \cite{efron2004}.

@@

Additionally, the degree of kurtosis and skewness also suggests that the normal distribution itself might not be a good approximation of the null. To address this issue, we also evaluated a mixture distribution of two normal components. Panel B of Fig~\ref{fig:1} demonstrates the goodness of fit of the mixture normal and the normal distribution using Kolmogorov-Smirnov (KS) test statistic computed on fitted normal and mixture normal distribution when fitted on CBEA scores in simulation scenarios under the global null. We can see that the mixture normal distribution is a better fit (lower KS scores) than the normal distribution across both sparsity and correlation settings.

@@

However, null distribution based on taxa-permutation is sensitive to inter-taxa correlations within the set \cite{wu2012}. Since the permutation procedure does not preserve correlation structures, estimating parameters from empirical scores on permuted data will underestimate the variance inflation due to correlation. We account for this by combining the mean estimate from permuted data with the variance estimate from unpermuted data, where the inter-taxa correlation structure remains undisturbed. However, this procedure assumes that the variance of CBEA is equal under both the null and alternate hypotheses.

\section*{Evaluation}
All code and data sets used for evaluation of this method is publicly available and can be found on GitHub (\href{www.github.com/qpmnguyen/CBEA\_analysis}{qpmnguyen/CBEA\_analysis}).

\subsection*{Simulation Analysis}
\subsubsection*{Data generation model}  
To address the performance of CBEA for different modeling tasks, we simulated microbiome count data under the assumption that it follows a zero-inflated negative binomial distribution, which is a good fit for real microbiome relative abundance data \cite{calgaro2020}. Suppose $X_{ij}$ are observed counts for a sample $i$ and taxon $j$, then we have the following probability model
\begin{equation}
    \mathbf{X}_{ij} =
      \begin{cases}
        0 & \text{with probability $p_j$}\\
        \mathbf{NB}(\mu_j, \phi_j) & \text{with probability $1 - p_j$}\\
      \end{cases}       
\end{equation}

where $\mu_j$ and $\phi_j$ are mean and dispersion parameters, respectively. To incorporate a flexible correlation structure into our simulation model, we utilized the NorTA (Normal to Anything) method \cite{cario1997}. Given an $n$ by $p$ matrix of values $\mathbf{U}$ sampled from multivariate normal distribution with correlation matrix $\mathbf{\rho}$, we can generate target microbiome count vector $\mathbf{X_{.j}}$ for taxa $j$ following the marginal distribution $\mathbf{NB}$ characterized by the negative binomial cumulative distribution function $\mathbb{F_{\mathbf{NB}}}$:
\begin{equation}
    \mathbf{X}_{.j} = \mathbb{F_{\mathbf{NB}}}^{-1}(\Phi_{U_i})
\end{equation}
In this instance, for each taxon $j$, we set elements in $\mathbf{U}_{.j}$ to be zero with probability $p_j$ and applied $\mathbf{NB}^{-1}(\mu_j, \phi_j)$ on non-zero elements to generate our final count matrix $\mathbf{X}$. To ensure that our simulations match closely to real data, we fitted negative binomial distribution using a maximum likelihood approach (with the \emph{fitdistrplus} package in R \cite{delignette-muller2015}) to non-zero counts for each taxon from 16S rRNA profiling of stool samples from the Human Microbiome Project (HMP). We take the median values of the estimated mean and dispersion parameters as the baseline of our simulations. For simplicity, we assumed that inter-taxa correlation follows an exchangeable structure

\subsubsection*{Simulation scenarios for enrichment analysis at the sample level}
To assess type I error rate and power for enrichment significance testing at the sample level, we simulated data based on the schema above, and assessed enrichment for one focal set. Type I error was obtained under the global null as the number of samples where the null hypothesis was rejected at $\alpha = 0.05$ over the total number of samples (which represents the total number of hypotheses tested). Power was obtained using the same formulation as type I error rate but under the global alternate. We treated type I error and power as estimates of binomial proportions and utilized the Agresti-Couli \cite{agresti1998} formulation to calculate 95\% confidence intervals. Across both analyses, we varied sparsity levels ($p = 0.2, 0.4, 0.6$) and inter-taxa correlation within the set ($\rho = 0, 0.2, 0.5$). For type I error analysis, we also varied the size of the set (50, 100, 150). For power analyses, set size was kept constant at 100 but different effect sizes (fold change of 1.5, 2, and 3). All sample sizes were set at 10,000.

For classifiability, we evaluated the scores against the true labels per sample (indicating the sample has a set with inflated counts) using the area under the receiving operator curve (AUC). This is a strategy used in Frost \cite{frost2020} which evaluates the informativeness of scores by assessing the relative ranking of samples (i.e. whether samples with inflated counts are highly ranked using estimated scores).  DeLong 95\% confidence intervals for AUC \cite{delong1988} were obtained for each estimate. Simulation settings for classification performance were identical to power analyses as detailed in the previous paragraph.

\subsubsection*{Simulation scenarios for enrichment analysis at the population level}
To assess type I error rate and power for differential abundance testing task, we simulated data based on the schema above, and assessed differential abundance of 50 sets with 100 taxa per set across 20 replicates per simulation condition. Type I error is calculated as the number of differentially abundant sets over the total number of sets for each simulation under the global null. Power is defined similarly, but instead under the global alternate hypothesis. Estimates and confidence intervals for type I error and power are calculated as cross-replicate mean and standard error. A set is differentially abundant when all taxa within a set are differentially abundant with the same effect size. Across both analyses, we varied sparsity levels ($p = 0.2, 0.4, 0.6$), and inter-taxa correlation within the set ($\rho = 0, 0.2, 0.5$). Half of the sets are differentially abundant across case/control status with varying effect sizes (fold change of 1.5, 2, and 3). Due to the compositional nature of microbiome taxonomic data, simple inflation of raw counts would cause an artificial decrease in the abundance of the remaining un-inflated sets. As such, we applied a compensation procedure as described in Hawinkel et al. \cite{hawinkel2019} to ensure the validity of simulation results. All sample sizes were set at 2,000.

\subsubsection*{Simulation scenarios for downstream prediction}
To assess predictive performance, we generated predictors based on the simulation schema presented above and evaluated prediction for both binary and continuous outcomes using a standard random forest model \cite{breiman2001}. For binary outcomes, we use AUC similar to the classification analyses above. For continuous outcomes, we used root mean squared error (RMSE). All predictive model fitting was performed using \emph{tidymodels} \cite{kuhn2020b} suite of packages. Across both learning tasks, we varied sparsity ($p = 0.2, 0.4, 0.6$), and inter-taxa correlation ($\rho = 0, 0.2, 0.5$). Continuous outcomes $Y_{cont}$ were generated as linear combinations of taxa counts.  
\begin{equation}
    Y_{cont} = f(\mathbf{X}) + \mathbf{\epsilon}
\end{equation}
where $\mathbf{\epsilon} \sim N(0, \sigma_{\epsilon}^2)$ and $f(\mathbf{X}) = \beta_0 + \mathbf{X}\mathbf{\beta}$. For each simulation, we set $\beta_0$ to be $\frac{6}{\sqrt{10}}$ similar to \cite{xiao2018}. The degree of model saturation (the number of non zero $\mathbf{\beta}$ values) were varied between 0.1 and 0.5, and signal to noise ratio (SNR = $\frac{\sigma(f(\mathbf{X}))}{\sigma_{\epsilon}}$) was varied between 1.5, 2, and 3.

For binary outcomes, we generate $Y_{binary}$ as Bernoulli draws with probability $p_{binary}$, where 
\begin{equation}
    p_{binary} = \frac{1}{1 + \exp(f(\mathbf{X}) + \mathbf{\epsilon})}
\end{equation}
To ensure a balance of classes, we applied the strategy described in Dong et al. \cite{dong2020} where the associated $\beta$ values are evenly split between positive and negative associations. All data sets generated from prediction tasks have 2,000 samples with 5,000 taxa over 50 sets with a size of 100 taxa per set.

\subsection*{Evaluation using real data}
In addition to simulation analyses, we also evaluated our method using real data sets based on both 16S rRNA gene sequencing and whole-genome sequencing. We follow the benchmark set by Geistlinger et al. \cite{geistlinger2021} All data sets are obtained from either the \emph{curatedMetagenomicData} \cite{pasolli2017} and \emph{HMP16SData} \cite{schiffer2019} R packages (2020-10-02 snapshot), or downloaded from the Qiita platform \cite{gonzalez2018}.

\noindent \textbf{Single Sample Enrichment}: To assess the false discovery rate and true discovery rate of CBEA in sample-level enrichment testing, we utilized the 16S rRNA gene sequencing of the oral microbiome at the gingival subsite from the Human Microbiome Project \cite{consortium2012, proctor2019}. We utilized this data set following the approach outlined in Calagaro et al. \cite{calgaro2020}. This data set is approximately labeled, where aerobic microbes are enriched in the supragingival subsite where the biofilm is exposed to the open air, while conversely anaerobic microbes thrive in the subgingival site \cite{thurnheer2016}. Here, we assessed the enrichment of aerobic microbes across all samples, we considered the false positive rate as the number of samples from the subgingival site with significant enrichment, and the true positive rate as the number of supragingival samples with significant enrichment. Microbial tropism annotation at the genus level was from Beghini et al. \cite{beghini2019} and was downloaded directly from the GitHub repository associated with Calagaro et al. \cite{matteocalgaro2020}.

\noindent \textbf{Differential Abundance Analsysis}: To assess type I error using CBEA scores in differential abundance analysis, we utilized the 16S rRNA gene sequencing of stool samples from the Human Microbiome Project \cite{consortium2012, proctor2019}. Here, we randomly assign samples a label of case or control, and repeated this process 500 times, assessing all candidate methods at each iteration. Type I error is then the number of taxa identified as differentially abundant across all tested taxa. For the true positive rate, we used the same gingival data set as described above. However, instead of testing for aerobic microbes as a group, the true positive rate is the number of aerobic/anaerobic genera identified as differentially abundant across all aerobic or anaerobic genera.

\noindent \textbf{Disease Prediction}: To assess predictive power, we utilized the whole genome sequencing of stool samples of inflammatory bowel disease (IBD) patients from the MetaHIT consortium \cite{nielsen2014}. This data set contains 396 samples from a cohort of European adults, where 195 adults were classified as having IBD (which includes patients diagnosed with either ulcerative colitis or Crohn's disease). Additionally, we also utilized a similar data set from Gevers et al. \cite{gevers2014} which also profiles the gut microbiome of IBD patients and controls but using 16S rRNA gene sequencing. This data set contains 16S rRNA gene sequencing samples from a cohort of pediatric patients (ages $<$ 17) from the RISK cohort enrolled in the United States and Canada. Of the 671 samples obtained, 500 samples belong to patients with IBD.

\subsubsection*{Comparison Methods}

\noindent \textbf{Single sample enrichment}: For type I error and power analyses, we compared the CBEA method with a naive Wilcoxon rank sum test. We added a pseudocount of 1 to all values. This is a non-parametric difference in means test, where we compared the abundance of taxa of a pre-defined set and its complement within a single sample. For classification performance, we compared CBEA methods against GSVA \cite{hanzelmann2013}, ssGSEA \cite{barbie2009}, and the W-statistic from the Wilcoxon rank sum test. All three approaches were applied directly on count data (after pseudocount). For GSVA, the Poisson kernel was used.

\noindent \textbf{Differential Abundance}: Since CBEA are sample-level enrichment scores, we performed differential abundance by using a Wilcoxon Rank Sum test and Welch's t-test across case/control status on CBEA generated scores. We added a pseudocount of 1 to all values. For comparison, we chose representative state-of-the-art methods in differential abundance analysis, namely DESeq2 \cite{love2014,mcmurdie2014} and corncob \cite{martin2020}. For DESeq2, we performed a likelihood ratio test against an intercept only reduced model with dispersion estimated with local fit. For corncob, we also performed a likelihood ratio test against an intercept only reduced model without bootstrapping.

\noindent \textbf{Disease Prediction}: We fit random forest on CBEA scores, as well as ssGSEA \cite{hanzelmann2013} and GSVA \cite{barbie2009} similar to single sample enrichment section. We added a pseudocount of 1 to all values. Additionally, we also compared performance using CBEA against a standard analysis plan where the centered log-ratio transformation (CLR) was applied to count-aggregated sets as inputs to a machine learning model.

\section*{Results}
In this section, we present the performance of our proposed method for three applicable microbiome analysis tasks: sample level enrichment, differential abundance, and disease prediction. We obtained these results from both parametric simulations and examples from real data.

\subsection*{Enrichment testing at the sample level}
CBEA provides significance testing for enrichment at the sample level using the null distribution estimation procedure described in \nameref{methods}. Here, we present empirical results for this application of CBEA assessing type I error, power, and classification capacity.

\subsubsection*{Simulation studies}
Panel A and B in Fig~\ref{fig:2} demonstrate type I error and power respectively across different simulation conditions. We benchmarked the results of the CBEA method against a naive Wilcoxon rank sum test performed at the sample level, comparing the mean count difference between taxa in the set its complement. All methods demonstrate good type I error control at $\alpha = 0.05$ under zero correlation across all simulation conditions. However, under both medium ($\rho = 0.2$) and high ($\rho = 0.5$) correlation settings, both the Wilcoxon test and unadjusted CBEA variants show high levels of inflated type I error, where Wilcoxon test performed the worst. On the other hand, adjusted CBEA methods (under both distributions) control for type I error at the appropriate $\alpha$ level even at high correlations.

\begin{figure}[!h]
    \centering
    \includegraphics[width=\textwidth]{figures/sim_data_ss_hypo.png}
    \caption{Sample-level inference with CBEA under parametric simulations (\textbf{(A)} and \textbf{B}), and real data analysis (\textbf{(C)} and \textbf{(D)}). In simulation analyses, panel \textbf{(A)} shows type I error rate, and panel \textbf{(B)} shows power for single sample enrichment test for a specified set and was compared against a Wilcoxon rank sum test at $\alpha$ of 0.05. In real data analysis, panel \textbf{(C)} shows the false-positive rate, and panel \textbf{(D)} shows the true positive rate. For this analysis, 16S rRNA data from the oral microbiome of the gingival site was used. The set of aerobic microbes was tested for enrichment in all samples and was identified as correctly enriched if a significant $p$-value was obtained in supragingival samples. Confidence bounds were obtained using Agresti-Couli \cite{agresti1998} approach. Adjusted CBEA demonstrated control of type I error at the appropriate $\alpha$ level while remaining methods (not included in subsequent power analyses) showed an inflated type I error rate. However, this resulted in lower power for adjusted CBEA methods.}
    \label{fig:2}
\end{figure}

@@

\subsection*{Disease Prediction}   
Since CBEA can generate informative scores that can discriminate between samples with inflated counts for a set (Fig~\ref{fig:2}), we want to assess whether they can also act as useful inputs to predictive models. In this section we assessed the predictive performance of a naive random forest model \cite{breiman2001} with different single sample enrichment scoring methods as inputs (evaluating CBEA, ssGSEA, and GSVA). Additionally, we also compared predictive performance of using these scores against the a standard approach of using the centered log ratio transformation (CLR) on taxon sets aggregated via abundance summations.     
\subsubsection*{Simulation studies}
Fig~\ref{fig:6} shows results for simulation studies as detailed in the \nameref{methods} section. Panel A presents results for a regression learning task with a continuous outcome while panel B presents results for a classification task with a binary outcome. As expected, performance across all assessed methods increased with a higher signal-to-noise ratio. Both CLR and CBEA approaches outperformed both GSVA and ssGSEA across all simulation conditions and learning tasks. This is because both GSVA and ssGSEA are more sensitive to the degree of inter-taxa correlation and sparsity, while CBEA and CLR did not experience a similar level of impact. As such, performance gap widens with increasing correlation and sparsity. Interestingly, this difference in performance is not as pronounced under high levels of effect saturation (across both learning tasks), suggesting that when there is a high number of sets contributing to an effect, model choice might not be as important.

@@

These results demonstrate that CBEA generated scores are informative features in disease prediction tasks. Simulation results indicate that CBEA methods perform much better than either GSVA or ssGSEA, but not as well as the standard CLR approach. Interestingly, however, CBEA methods were much more competitive with CLR in either WGS data sets or data sets with higher sparsity levels.

>>>>

% Adding overline macro 
\newcommand{\overbar}[1]{\mkern 2mu\overline{\mkern-2mu#1\mkern-20mu}\mkern 20mu}

@@

However, there exists unique challenges in the analysis of these taxonomic count tables \cite{li2019a,li2015}. The data is sparse, high-dimensional, and likely compositional \cite{gloor2017, li2019a, li2015}. Even though the these problems are challenging, a very approachable solution is to to use set-based analysis methods, also termed gene set testing in the genomics literature \cite{khatri2012, goeman2007}. Aggregated variables can be less sparse, and testing on a smaller number of features can reduce the multiple-testing burden. As such, gene set testing methods have been shown to increase power and reproducibility of genomic analyses. Furthermore, through the usage of functionally informative sets defined apriori based on historical experiments (for example MSigDB \cite{subramanian2005}, and Gene Ontology \cite{ashburner2000}), gene set analysis also allows for more biologically informative interpretations.

There exists a diverse set of methods already developed in this field. Traditional methods utilize the hypergeometric distribution to test for the overrepresentation of a gene set using a candidate list of genes screened based on a marginal model \cite{goeman2007}. Unfortunately, these approaches are sensitive to the differential expression test as well as the chosen threshold when trying to select genes for the candidate list. Aggregate score methods, which are generally more preferred \cite{irizarry2009}, instead assigns a score for each gene set based on gene-specific statistics such as z-scores or fold change. Of these approaches, methods such as GSEA \cite{subramanian2005} performs a test for each gene set at the population level, summarizing information across all samples. Conversely, methods such as GSVA \cite{hanzelmann2013} and VAM \cite{frost2020}, generate enrichment scores at the sample level and are more akin to a transformation. In addition to being able to screen for enriched sets per sample, this strategy also allows for the flexible incorporation of different downstream analyses, such as fitting prediction models, or performing dimension reduction.

@@

In microbiome research, even though no explicit enrichment analysis is performed, researchers often aggregate taxa to higher Linnean classification levels such as genus, family, or phylum. However, despite this interest, there is limited research done to extend existing set-based methods to microbiome relative abundance data. Some software suites, such as \emph{MicrobiomeAnalyst}, do offer tools to perform enrichment testing with curated taxon sets \cite{chong2020}. However, the approach used in \emph{MicrobiomeAnalyst} is a form of overrepresentation analysis at the population level and therefore similarly sensitive to the differential abundance approach used and p-value threshold. One of the primary challenges for adapting gene set analysis to the microbiome context is the compositional nature of the data. Sequencing technologies constrain the total number of reads, and samples are expected to have the same number of reads instead of DNA content \cite{quinn2019, quinn2018}. However, different samples still yield arbitrarily different total read counts \cite{morton2019, gloor2017}, suggesting the use of normalization methods to allow for proper comparison of feature abundances across samples. However, microbiome data sets do not follow certain assumptions that enable the cross-application of methods from similar fields (such as RNA-seq) \cite{quinn2019, quinn2018}. For example, DESeq2's \emph{estimateSizeFactors} \cite{love2014} assumes that the majority of genes acts as housekeeping genes with constant expression levels across samples. As such, practitioners often rely on total sum normalization to transform count data into relative proportions that sum to one \cite{weiss2017}. Empirical studies have supported this decision, showing increased performance of analysis tasks under this normalization schema \cite{mcknight2019}. Since this approach imposes a sum constraint on the data, post normalization microbiome data sets are compositional \cite{gloor2017}, which means that the abundance of any taxa can only be interpreted relative to another. Under this scenario, log-ratio based approaches from the compositional data analysis (CoDA) literature \cite{aitchison1994} are motivated to address this issue.

@@

The CBEA method generates sample-specific enrichment scores for microbial sets using products of proportions \cite{egozcue2003}. Details on the computational implementation of CBEA can be found in the Supplementary Materials. The CBEA method takes two inputs:  

@@

This assumption is also supported by preliminary simulation studies (detailed below). In panel A of Fig~\ref{fig:1}, we simulated microbiome taxonomic count data under the global null across different data features and compute raw CBEA scores and compute kurtosis and skewness. It can be seen that the characteristics of the null change depending on sparsity and inter-taxa correlation. Sparsity seems to drive the distribution to be more positively skewed while inter-taxa correlation encourages platykurtic (negative kurtosis). The effect is most dramatic under both high inter-taxa correlation and sparsity. This heterogeneity further supports the decision to estimate an empirical null distribution, as suggested by Efron \cite{efron2004}.

@@

Additionally, the degree of kurtosis and skewness also suggests that the normal distribution itself might not be a good approximation of the null. To address this issue, we also evaluated a two-component normal mixture distribution. Panel B of Fig~\ref{fig:1} demonstrates the goodness of fit of the mixture normal and the normal distribution using Kolmogorov-Smirnov (KS) test statistic computed on fitted normal and mixture normal distribution when fitted on CBEA scores in simulation scenarios under the global null. We can see that the mixture normal distribution is a better fit (lower KS scores) than the normal distribution across both sparsity and correlation settings.

@@

\subsubsection*{Variance inflation due to inter-taxa correlation}  
When taxa within a set are highly correlated, the variance of the sample mean of taxon-wise statistics is inflated. Without loss of generalizability, for a set of taxa with taxa-specific statistics $x_1, ..., x_p$ we have the variance of the mean $\bar{x}$ to be:  
\begin{equation} \label{vareq:1}
    Var(\bar{x}) = \frac{1}{m^2}\left(\sum_{i = 1}(\sigma_i^2) + \sum_{i < j}\rho_{ij}\sigma_i\sigma_j\right)
\end{equation}
where $\sigma_i$ is the standard deviation of taxon $i$ while $\rho_{ij}$ is the correlation between $i$ and $j$. The second term of (\ref{vareq:1}) is the correlation dependent variance component, which goes to 0 if there is no correlation. The CBEA statistic follows a similar pattern. Since the geometric mean of a set of variables is equivalent to the exponential of the arithmetic mean of their logarithms, we can re-write CBEA score for a set $k$ with size $K$ as follows:  
\begin{equation}\label{vareq:2}
    M_{i,k} = \sqrt{\frac{K(p - K)}{K + (p - K)}} \left( \overbar{\log{X_{i,j|j \in K}}} - \overbar{\log{X_{i,j|j \notin K}}} \right)   
\end{equation}
where $p$ is the overall number of taxa, $j$ is the index of a taxa and $K$ is the set of indices of taxa in set $k$. The CBEA statistic then looks similar to a t-statistic for difference in means of log-transformed proportions. As such, the pooled variance of cILR is dependent on the variance inflation of both mean components $\overbar{\log{X_{i,j|j \in K}}}$ and $\overbar{\log{X_{i,j|j \notin K}}}$. The result of this variance inflation is inflated type I error since highly correlated sets are also detected as significantly enriched.

However, as Wu et al. \cite{wu2012} showed, performing column permutation to estimate the null distribution of a competitive test statistic doesn't allow for adequate capture of this variance inflation factor since the permutation procedure disrupts the natural correlation structure of the original variables. It is important to address this problem since there is strong inter-taxa correlation within the microbiome \cite{kurtz2015a}. Our strategy for addressing issue is location (or mean) estimate from the column permuted raw score matrix with the spread (or variance) estimate from the original un-permuted scores. This still allows us to leverage the null generated via column permutation while using the proper variance estimate taken from scores where the correlation structure has not been disrupted. As such, this procedure assumes that the variance of the test statistic under the alternate hypothesis is the same as that of the null. Details of the computational implementation to this estimation process can be found in the Supplementary Materials.

However, set-based analysis is an exploratory approach that can help generate functionally informative hypotheses, and as such users might not want strict type I error control in favor of higher power. This is especially true for competitive hypotheses, where its stricter formulation compared to the self-contained approach implies that the test naturally has lower power \cite{goeman2007, ackerman2009}. Furthermore, sets that are highly correlated compared to background can be biologically relevant. Therefore, CBEA provides an option for users to specify whether correlation adjustment is desired.

\section*{Evaluation}
We based our evaluation strategy on gene set testing benchmarking standards set by Geistlinger et al. \cite{geistlinger2021} and utilized the same approaches whenever possible. All data sets are obtained from either the \emph{curatedMetagenomicData} \cite{pasolli2017} and \emph{HMP16SData} \cite{schiffer2019} R packages (2020-10-02 snapshot), or downloaded from the Qiita platform \cite{gonzalez2018}. All code and data sets used for evaluation of this method is publicly available and can be found on GitHub (\href{www.github.com/qpmnguyen/CBEA\_analysis}{qpmnguyen/CBEA\_analysis}).

\subsection*{Statistical significance}
We evaluate the inference procedure of CBEA compared to alternate methods using two approaches: randomly sampled taxa sets and sample label permutation. These analyses were performed on the 16S rRNA gene sequencing of the oral microbiome from the Human Microbiome Project \cite{consortium2012, proctor2019}. This data set contains 369 samples split into two subsites: supragingival and subgingival.

\subsubsection*{Sample-level inference} 
Due to CBEA's self-contained null hypothesis, we can perform inference at the sample level for the enrichment of a set. We evaluated this application by generating one random taxa of different sizes $S \in \{20, 50, 100, 150, 200\}$ across 500 iterations. Random sets can act as our estimate for type I error since this matches the CBEA null hypothesis stated in \nameref{methods}, where we expect within each sample sets of randomly sampled taxa should not be significantly enriched compared to the remainder background taxa. For this evaluation, we estimated type I error as the fraction of samples that detects our random set as significant at a p-value threshold of 0.05 with confidence bands computed from the standard error across all iterations.  Additionally, this analysis also demonstrate whether CBEA is sensitive to different set sizes.

\subsubsection*{Population-level inference} 
We can perform enrichment testing at the population level by generating corresponding sample level CBEA scores and performing a two-sample test such as Welch's t-test. In order to evaluate CBEA under this context, we generated CBEA scores of sets representing genus-level annotation in above gingival data set \cite{consortium2012, proctor2019} and applied a t-test to test for enrichment (similar to GSVA \cite{hanzelmann2013}) across a randomly generated variable indicating case/control status (repeated 500 times). Type I error is estimated as the fraction of sets per iteration found to be significantly enriched with confidence bands computed from the standard error across all iterations. In addition, we also performed a random set analysis assessing, where we generated 100 sets of different set sizes $S \in \{20, 50, 100, 150, 200\}$ and evaluated the fraction of genera that were found to be differentially abundant across the original labels (supragingival versus subgingival subsite). 95\% confidence intervals were computed using the Agresti-Couli approach \cite{agresti1998}.

\subsection*{Phenotype relevance}
We want to evaluate whether sets found to be significantly enriched by CBEA are relevant to the research question. To perform this assessment, we relied on the gingival data set mentioned prior \cite{consortium2012, proctor2019}. This data set was chosen due to its clear biological interpretation that can serve as the ground truth. Specifically, we expect aerobic microbes to be enriched in the supragingival subsite where the biofilm is exposed to the open air, while conversely anaerobic microbes thrive in the subgingival site \cite{thurnheer2016}. Genus-level annotations for microbial metabolism from Beghini et al. \cite{beghini2019} and was obtained from the GitHub repository associated with Calagaro et al. \cite{matteocalgaro2020}. For sample-level inference, we assessed power as the fraction of supragingival samples where aerobic microbes are significantly enriched. For population-level inference, power is the fraction of sets representing genus level taxonomic assignments were significant across subsite labels.

In addition to statistical power, we also assessed phenotype relevance through evaluating whether highly ranked sets based on CBEA scores are more likely to be enriched according to ground truth. This is represented by the area under the receiving operator curve (AUROC/AUC) scores computed on CBEA scores against true labels (similar to strategy employed by VAM \cite{frost2020}). DeLong 95\% confidence intervals for AUC \cite{delong1988} were obtained for each estimate.

\subsection*{Disease Prediction}
CBEA scores can also be used downstream analyses such as disease prediction tasks. We utilized two data sets for this evaluation:

\begin{enumerate}
\item Whole genome sequencing of stool samples from inflammatory bowel disease (IBD) patients in the MetaHIT consortium \cite{nielsen2014}. This data set contains 396 samples from a cohort of European adults, where 195 adults were classified as having IBD (which includes patients diagnosed with either ulcerative colitis or Crohn's disease).

\item 16S rRNA gene sequencing of stool samples from IBD patients in the pediatric RISK cohort \cite{gevers2014}. This data set contains 16S rRNA gene sequencing samples from a cohort of pediatric patients (ages $<$ 17) from the RISK cohort enrolled in the United States and Canada. Of the 671 samples obtained, 500 samples belong to patients with IBD.   
\end{enumerate}

We evaluate disease prediction performance by fitting a random forest model \cite{breiman2001} using inputs as CBEA scores to classify samples of patients with IBD and healthy controls. Random forest was chosen as a baseline learner due to its flexibility as an out-of-the-box model that is easy to fit without requiring complex hyperparameter tuning.

\subsection*{Comparison Methods} 
We benchmarked the statistical properties of CBEA against existing baseline approaches. For sample-level inference analyses, utilized the Wilcoxon rank-sum test, which non-parametrically tests the difference in mean counts between taxa from a pre-defined set and its remainder similar to CBEA. For assessments at the population level, we compared CBEA against performing a standard test for differential abundance with set-level features generated via element-wise summations instead. We chose DESeq2 \cite{love2014} and corncob \cite{martin2020} since they represent both methods extrapolated from RNA-seq \cite{mcmurdie2014} and those developed specifically for microbiome data.

Since disease prediction models and rankings-based phenotype relevance analyses seek to evaluate the informativeness of CBEA scores instead of relying on computing p-values, we compared performance against other single sample based approaches from the gene set testing literature, specifically ssGSEA \cite{barbie2009} and GSVA \cite{hanzelmann2013}. Additionally, for evaluating prediction, we also compared performance against a standard analysis plan where inputs are count-aggregated sets with the centered log-ratio (CLR) transformation.

\section*{Results}
In this section, we present results for evaluating statistical significance, phenotype relevance, and predictive performance. In addition to real data, we also evaluated models based on parametric simulations. Results for this section can be found in Supplemental Materials.

\subsection*{Sample level inference}
CBEA provides significance testing for enrichment at the sample level using the null distribution estimation procedure described in \nameref{methods}. Here, we present empirical results for this application of CBEA assessing type I error, power, and classification capacity.

@@

\subsection*{Disease Prediction}   
Since CBEA can generate informative scores that can discriminate between samples with inflated counts for a set (Fig~\ref{fig:2}), we want to assess whether they can also act as useful inputs to predictive models. In this section we assessed the predictive performance of a naive random forest model \cite{breiman2001} with different single sample enrichment scoring methods as inputs (evaluating CBEA, ssGSEA, and GSVA). Additionally, we also compared predictive performance of using these scores against the a standard approach of using the centered log ratio transformation (CLR) on taxon sets aggregated via abundance summations.

\subsubsection*{Simulation studies}
Fig~\ref{fig:6} shows results for simulation studies as detailed in the \nameref{methods} section. Panel A presents results for a regression learning task with a continuous outcome while panel B presents results for a classification task with a binary outcome. As expected, performance across all assessed methods increased with a higher signal-to-noise ratio. Both CLR and CBEA approaches outperformed both GSVA and ssGSEA across all simulation conditions and learning tasks. This is because both GSVA and ssGSEA are more sensitive to the degree of inter-taxa correlation and sparsity, while CBEA and CLR did not experience a similar level of impact. As such, performance gap widens with increasing correlation and sparsity. Interestingly, this difference in performance is not as pronounced under high levels of effect saturation (across both learning tasks), suggesting that when there is a high number of sets contributing to an effect, model choice might not be as important.

@@

These results demonstrate for given set-based features, CBEA can generate scores can could be informative in disease prediction tasks. Simulation results indicate that CBEA methods perform much better than either GSVA or ssGSEA, but not as well as the standard CLR approach. Interestingly, however, CBEA methods were much more competitive with CLR in either WGS data sets or data sets with higher sparsity levels.
